Notes on categorical logic and term calculi

### 2025-10-02 term-calculus/cat_logic_notes.tex
wip

\[
\inferrule*[right=$id$]{\vdash A\;\text{type}}{x : A \vdash x: A}
\]
\[
\inferrule*[right=$fI$]{f \in \mathcal{G}(A_1, \ldots, A_n; B)\;\Gamma_1 \vdash M_1 : A_1, \ldots, \Gamma_n \vdash M_n : A_n\;\text{Shuf}(\Gamma_1,\ldots,\Gamma_n;\Phi)}{\Phi \vdash f(M_1, \ldots, M_n) : B}
\]
\[
\inferrule*[right=$\otimes I$]{\Gamma \vdash N : A \qquad \Delta \vdash M : B \qquad \text{Shuf}(\Gamma,\Delta;\Phi)}{\Phi \vdash \{N \otimes M\} : A \otimes B}
\]
\textcolor{red}{The rule above seems to imply the tensor is not strict.}
\textcolor{red}{Can the below rule be made admissible? It looks like a composition.}
\[
\inferrule*[right=$\otimes E$]{\Psi \vdash M : A \otimes B \quad \Gamma, x : A, y : B \vdash N : C \quad \text{Shuf}(\Gamma,\Psi;\Phi)}{\Phi \vdash \text{match}_{A \otimes B}(M, xy.N) : C}
\]
\[
\inferrule*[right=$\mathbf{1}I$]{\;}{() \vdash \star : \mathbf{1}} \qquad \inferrule*[right=$\mathbf{1}E$]{\Psi \vdash M : \mathbf{1} \qquad \Gamma,\Delta \vdash N : A}{\Gamma, \Psi. \Delta \vdash \text{match}_{\mathbf{1}}(M,N) : A}
\]
\[
\inferrule*[right=$\multimap I$]{\Gamma, x : A \vdash M : B}{\Gamma \vdash \lambda x . M : A \multimap B}
\]
\textcolor{red}{Should there be a symmetry involved in the rule above?}
\[
\inferrule*[right=$\multimap E$]{\Gamma \vdash M : A \multimap B \quad \Delta \vdash N : A \quad \text{Shuf}(\Gamma, \Delta; \Phi)}{\Phi \vdash M N : B}
\]

To implement the universal properties, we need the following equations for the tensor ($\beta$ and $\eta$ laws)
\[
\text{match}_{\otimes}(\{M,N\}, xy.P) \equiv P[M / x, N /y]
\]
\[
\text{match}_{\otimes}(M, xy.N[\{x,y\}/ u]) \equiv N[M / u]
\]

\[
\text{match}_{\mathbf{1}}(\star, N) \equiv N \qquad \text{match}_{\mathbf{1}}(M,N[\star / u]) \equiv N[M / u]
\]

@@

@@

\[
\inferrule*{
  x : A \vdash x : A \qquad () \vdash \star : \mathbf{1}
}{x : A \vdash \{x,\star\} : A \otimes \mathbf{1}}
\]
the morphism $A \otimes \mathbf{1} \to A$ is then
\[
\inferrule*{
\inferrule*{}{z : A \otimes \mathbf{1} \vdash z : A \otimes \mathbf{1}}
\qquad
\inferrule*{y : \mathbf{1} \vdash y : \mathbf{1} \qquad x : A \vdash x : A}{x : A, y : \mathbf{1} \vdash \text{match}_{\mathbf{1}}(y,x) : A}
}{z : A \otimes \mathbf{1} \vdash \text{match}_{A \otimes \mathbf{1}}(z, xy.\text{match}_{\mathbf{1}}(y,x)) : A}
\]
Then we have
\begin{itemize}
  \item \[
  \inferrule*{
  \inferrule*{
  \inferrule*{        
  \inferrule*{
            \inferrule*{
  w : A \vdash w : A \qquad () \vdash \star : \mathbf{1}
}{w : A \vdash \{w,\star\} : A \otimes \mathbf{1}}
\qquad
\inferrule*{
\inferrule*{}{z : A \otimes \mathbf{1} \vdash z : A \otimes \mathbf{1}}
\qquad
\inferrule*{y : \mathbf{1} \vdash y : \mathbf{1} \qquad x : A \vdash x : A}{x : A, y : \mathbf{1} \vdash \text{match}_{\mathbf{1}}(y,x) : A}
}{z : A \otimes \mathbf{1} \vdash \text{match}_{A \otimes \mathbf{1}}(z, xy.\text{match}_{\mathbf{1}}(y,x)) : A}
}{w : A \vdash \text{match}_{A \otimes \mathbf{1}}(z, xy.\text{match}_{\mathbf{1}}(y,x))[\{w, \star\}/z]}}{
w : A \vdash \text{match}_{A \otimes \mathbf{1}}(\{w, \star\}, xy.\text{match}_{\mathbf{1}}(y,x))}}{
w : A \vdash \text{match}_{\mathbf{1}}(y,x)[w / x,  \star / y]}}{
w : A \vdash \text{match}_{\mathbf{1}}(\star,w) \equiv w
}
      \]
  \item The other direction is substitution 
  \[
  \inferrule*{
  \inferrule*{z : A \otimes \mathbf{1} \vdash \{w, \star\}[\text{match}_{A \otimes \mathbf{1}}(z, xy.\text{match}_{\mathbf{1}}(y,x)) / w]}
{z : A \otimes \mathbf{1} \vdash \{\text{match}_{A \otimes \mathbf{1}}(z, xy.\text{match}_{\mathbf{1}}(y,x)), \star\}}}
{z : A \otimes \mathbf{1} \vdash \text{\textcolor{red}{What's next?}}}
  \]

@@

and prove the analogous statements, i.e. admissibility of substitution, associativity of substitution etc.
Then we can say that a free semilattice-enriched symmetric monoidal category is given by appropriate type theory.

\begin{example}
  Recall the raw morphism representation for our e-graph example
  \[(a \otimes ((2;\comonoid) \otimes (1;id));\sym);((* \otimes \counit + (id \otimes \counit \otimes id);<\!\!<) \otimes id);/~,\]
  We can render it using the  \enquote{enhanced} term syntax.
  \[
  \inferrule*{
  \inferrule*{
  \inferrule*{\vdash a : Int \quad \vdash 2 : Int}{\vdash a * 2}
  \quad
  \inferrule*{\vdash a : Int \quad \vdash 1 : Int}{\vdash a <\!\!< 1}
  }{\vdash (a * 2) + (a <\!\!< 1)}
  \quad
  \inferrule*{}{\vdash 1 : Int}
  }
  {((a * 2) + (a <\!\!< 1)) / 2}
  \]
\end{example}

>>>>

\[
\inferrule*[right=$id$]{\vdash A\;\text{type}}{x : A \vdash x: A}
\]
\[
\inferrule*[right=$fI$]{f \in \mathcal{G}(A_1, \ldots, A_n; B)\;\Gamma_1 \vdash M_1 : A_1, \ldots, \Gamma_n \vdash M_n : A_n\;\text{Shuf}(\Gamma_1,\ldots,\Gamma_n;\Phi)}{\Phi \vdash f(M_1, \ldots, M_n) : B}
\]
\[
\inferrule*[right=$\otimes I$]{\Gamma \vdash N : A \qquad \Delta \vdash M : B \qquad \text{Shuf}(\Gamma,\Delta;\Phi)}{\Phi \vdash \{N \otimes M\} : A \otimes B}
\]
% \textcolor{red}{The rule above seems to imply the tensor is not strict.}
% \textcolor{red}{Can the below rule be made admissible? It looks like a composition.}
\[
\inferrule*[right=$\otimes E$]{\Psi \vdash M : A \otimes B \quad \Gamma, x : A, y : B \vdash N : C \quad \text{Shuf}(\Gamma,\Psi;\Phi)}{\Phi \vdash \text{match}_{A \otimes B}(M, xy.N) : C}
\]
\[
\inferrule*[right=$\mathbf{1}I$]{\;}{() \vdash \star : \mathbf{1}} \qquad \inferrule*[right=$\mathbf{1}E$]{\Psi \vdash M : \mathbf{1} \qquad \Gamma,\Delta \vdash N : A}{\Gamma, \Psi. \Delta \vdash \text{match}_{\mathbf{1}}(M,N) : A}
\]
\[
\inferrule*[right=$\multimap I$]{\Gamma, x : A \vdash M : B}{\Gamma \vdash \lambda x . M : A \multimap B}
\]
% \textcolor{red}{Should there be a symmetry involved in the rule above?}
\[
\inferrule*[right=$\multimap E$]{\Gamma \vdash M : A \multimap B \quad \Delta \vdash N : A \quad \text{Shuf}(\Gamma, \Delta; \Phi)}{\Phi \vdash M N : B}
\]

To implement the universal properties, we need the following equations for the tensor ($\beta$ and $\eta$ laws)
\[
\text{match}_{\otimes}(\{M,N\}, xy.P) \equiv_{\beta_{\otimes}} P[M / x, N /y]
\]
\[
\text{match}_{\otimes}(M, xy.N[\{x,y\}/ u]) \equiv_{\eta_{\otimes}} N[M / u]
\]

\[
\text{match}_{\mathbf{1}}(\star, N) \equiv_{\beta_{\mathbf{1}}} N \qquad \text{match}_{\mathbf{1}}(M,N[\star / u]) \equiv_{\eta_{\mathbf{1}}} N[M / u]
\]

@@

\begin{lemma}

Each term paired with its context uniquely determines the derivation.
\end{lemma}

\begin{example}
Let's show that the term $f(g(h(x)))$ has a unique derivation.

\[
\inferrule*{
  \inferrule*{h \in \mathcal{G} \quad x : A \vdash x : A}{
  \inferrule*{g \in \mathcal{G} \quad x : A \vdash h(x) : B}{f \in \mathcal{G} \quad x : A \vdash g(h(x)) : C}}}
{x : A \vdash f(g(h(x))) : D}
\]

it might be tempting to write down another derivation like
\[
\inferrule*{f \in \mathcal{G} \quad y : B \vdash g(y) : C}{y : B \vdash f(g(y)) : D}
\]
but then you get stuck since the only way to apply this to $h(x)$ is by unraveling $f(g(y))$.
\end{example}

@@

\[
\inferrule*{
  x : A \vdash x : A \qquad () \vdash \star : \mathbf{1}
}{x : A \vdash \{x,\star\} : A \otimes \mathbf{1}}
\]
the morphism $A \otimes \mathbf{1} \to A$ is then
\[
\inferrule*{
\inferrule*{}{z : A \otimes \mathbf{1} \vdash z : A \otimes \mathbf{1}}
\qquad
\inferrule*{y : \mathbf{1} \vdash y : \mathbf{1} \qquad x : A \vdash x : A}{x : A, y : \mathbf{1} \vdash \text{match}_{\mathbf{1}}(y,x) : A}
}{z : A \otimes \mathbf{1} \vdash \text{match}_{A \otimes \mathbf{1}}(z, xy.\text{match}_{\mathbf{1}}(y,x)) : A}
\]
Then we have
\begin{itemize}
  \item \[
  \inferrule*{
            \inferrule*{
  w : A \vdash w : A \qquad () \vdash \star : \mathbf{1}
}{w : A \vdash \{w,\star\} : A \otimes \mathbf{1}}
\qquad
\inferrule*{
\inferrule*{}{z : A \otimes \mathbf{1} \vdash z : A \otimes \mathbf{1}}
\qquad
\inferrule*{y : \mathbf{1} \vdash y : \mathbf{1} \qquad x : A \vdash x : A}{x : A, y : \mathbf{1} \vdash \text{match}_{\mathbf{1}}(y,x) : A}
}{z : A \otimes \mathbf{1} \vdash \text{match}_{A \otimes \mathbf{1}}(z, xy.\text{match}_{\mathbf{1}}(y,x)) : A}
}{w : A \vdash \text{match}_{A \otimes \mathbf{1}}(z, xy.\text{match}_{\mathbf{1}}(y,x))[\{w, \star\}/z]}
\]

\begin{align*}
  \text{match}_{A \otimes \mathbf{1}}(z, xy.\text{match}_{\mathbf{1}}(y,x))[\{w, \star\}/z] &\equiv \text{match}_{A \otimes \mathbf{1}}(\{w, \star\}, xy.\text{match}_{\mathbf{1}}(y,x))\\
  &\equiv_{\beta_{\otimes}} \text{match}_{\mathbf{1}}(\star,w)\\
  &\equiv_{\beta_{\mathbf{1}}} w
\end{align*}

  \item The other direction is substitution 
  \[
  \inferrule*{z : A \otimes \mathbf{1} \vdash \{w, \star\}[\text{match}_{A \otimes \mathbf{1}}(z, xy.\text{match}_{\mathbf{1}}(y,x)) / w]}
{z : A \otimes \mathbf{1} \vdash \{\text{match}_{A \otimes \mathbf{1}}(z, xy.\text{match}_{\mathbf{1}}(y,x)), \star\}}
  \]

\begin{align*}
  \underbrace{\{\text{match}_{A \otimes \mathbf{1}}(\underbrace{z}_{M}, xy.\text{match}_{\mathbf{1}}(y,x)), \star\}}_{N} &\equiv_{\eta_{\otimes}} \text{match}_{A \otimes \mathbf{1}}(z, xy.\{\text{match}_{A \otimes \mathbf{1}}(\{x,y\}, x'y'.\text{match}_{\mathbf{1}}(y',x')), \star\})\\ 
  &\equiv_{\beta_{\otimes}} \text{match}_{A \otimes \mathbf{1}}(z,xy.\{{\text{match}_{\mathbf{1}}(y,x), \star}\})\\
  &\equiv_{\eta_{\mathbf{1}}}\text{match}_{A \otimes \mathbf{1}}(z, xy.\text{match}_{\mathbf{1}}(y,\{\text{match}_{\mathbf{1}}(\star,x), \star\}))\\
  &\equiv_{\beta_{\mathbf{1}}} \text{match}_{A \otimes \mathbf{1}}(z, xy.\text{match}_{\mathbf{1}}(\underbrace{y}_{M},\underbrace{\{x, \star\}}_{N[\star / u]}))\\
  &\equiv_{\beta_{\mathbf{1}}} \text{match}_{A \otimes \mathbf{1}}(z, xy.\{x, y\})\\
  &\equiv_{\eta_{\otimes}} z\\
\end{align*}

The above proof uses a more general statement that 
\begin{align*}
\{\text{match}_{\otimes}(M, xy.N), P\} &\equiv_{\eta_{\otimes}} \text{match}_{\otimes}(M,xy.\{\text{match}_{\otimes}(\{x,y\}, x'y'.N), P\})\\
&\equiv_{\beta_{\otimes}} \text{match}(M,xy.\{N[x/x',y/y'], P\})\\
&\equiv_{\alpha} \text{match}_{\otimes}(M,xy.\{N,P\})
\end{align*}

\textcolor{red}{What's the intuition behind these rules?}

@@

We can then aim to prove the analogous statements, i.e. admissibility of substitution, associativity of substitution etc.
Then we can say that a free semilattice-enriched symmetric monoidal category is given by appropriate type theory.

We will have the following equations
\begin{align*}
  \{N_1 + \ldots + N_n, M\} &\equiv \{N_1, M\} + \ldots + \{N_n, M\}\\
  \{M, N_1, \ldots, N_n\} &\equiv \{M, N_1\} + \ldots + \{M, N_n\}\\
  N_1 + \ldots + (M_1 + \ldots + M_m)  + \ldots + N_n &\equiv N_1 + \ldots + M_1 + \ldots + M_m + \ldots + N_n\\
  M + N &\equiv N + M\\
  N &\equiv N + N
\end{align*}

and the substitution is (it should follow from admissibility)
\begin{align*}
(N_1 + \ldots + N_n)[M/x] &= (N_1[M/x] + \ldots + N_n[M/x])\\
N[(M_1 + \ldots + M_n)/x] &= N[M_1/x] + \ldots + N[M_n/x]
\end{align*}

\begin{lemma}[Substitution is admissible]
  If $\Gamma \vdash M : A$ and $\Delta, x : A, \Phi \vdash N$ are derivable, then $\Delta, \Gamma, \Phi \vdash N[M/x]$ is derivable.
\end{lemma}
\begin{proof}
  Since the substitution is admissible in the original type theory, we only need to consider the extra rule that we introduced.
  Suppose $N \Coloneqq N_1 + \ldots + N_n$ and therefore comes from the rule
  \[
  \inferrule[]{\inferrule[]{\vdots}{\Gamma \vdash M} \quad \inferrule[]{\Delta, x : A, \Phi \vdash N_1, \ldots, \Delta, x : A, \Phi \vdash N_n}{\Delta, x : A, \Phi \vdash N}}{\;}
  \]
  By IH we have derivations
  \[
    \Delta, \Gamma, \Phi \vdash N_1[M/x], \ldots, \Delta, \Gamma, \Phi \vdash N_n[M/x]
  \]
  and therefore can get a derivation 
  \[
  \Delta, \Gamma, \Phi \vdash N_1[M/x] + \ldots + N_n[M/x]
  \]

  The second rule is actually is not given by the lemma above, so we need to build it into the rules \textcolor{red}{How? I think if I extend every rule with a $+$ in the premise, it will absorb the rules above.}
  % \[
  % \inferrule[]{\Gamma \vdash N_1 + \ldots + N_n \quad f \in \mathcal{G}}{f(N_1) + \ldots + f(N_n)}
  % \]
  % then we have the following case.

\end{proof}

Consider the revised rules
\begin{align*}
  \inferrule[]{\vdash A\;\text{type}}{x : A \vdash x + \ldots + x : A}\\
  \inferrule[]{f \in G \quad \Gamma_1 \vdash M_{11} + \ldots + M_{1 i_1} \ldots \Gamma_n \vdash M_{n 1} + \ldots + M_{n i_n}}{\Phi \vdash f(M_{1 1}, \ldots M_{n 1}) + \ldots + f(M_{1 i_1}, \ldots, M_{n i_n})}
\end{align*}

Then we can merge the substitution rules into a single rule
\[
(N_1 + \ldots + N_n)[M_1 + \ldots + M_m / x] = N_1[M_1 / x] + \ldots + N_n[M_1 / x] + \ldots + N_1[M_m / x] + \ldots + N_n[M_m / x]
\]

\begin{example}
  Recall the raw morphism representation for our e-graph example
  \[(a \otimes ((2;\comonoid) \otimes (1;id));\sym);((* \otimes \counit + (id \otimes \counit \otimes id);<\!\!<) \otimes id);/~,\]
  We can render it using the  \enquote{enhanced} term syntax.
  \[
  \inferrule*{
  \inferrule*{
  \inferrule*{\vdash a : Int \quad \vdash 2 : Int}{\vdash a * 2}
  \quad
  \inferrule*{\vdash a : Int \quad \vdash 1 : Int}{\vdash a <\!\!< 1}
  }{\vdash (a * 2) + (a <\!\!< 1)}
  \quad
  \inferrule*{}{\vdash 2 : Int}
  }
  {\vdash ((a * 2) + (a <\!\!< 1)) / 2}
  \]
  Recall the string diagram
  \[
  \tikzfig{./figures/e-graph-example-string}
  \]
  We can linerise it as
  \begin{minted}{haskell}
    let (x_1, x_2, x_3) = (a, 2, 1) in
      let y_1 = {(x_1 * x_2), (x_1 << x_3)} in
        y1 / x_2
  \end{minted}
  \textcolor{red}{Why would one be preferable over the other? One syntax provides a reasoning framework about monoidal e-graphs while the other is just a linearisation of hypergraphs? Equality procedure for hypergraphs?}
\end{example}
